\subsection{Error handling}
As the program reads the input file, the parser requests tokens from the scanner, who splits the
input text into tokens and relays them back to the parser. If an error is found, a message is
sent to \texttt{stderr} indicating the position (line and column) in the text where it was found,
and whether it was detected by the scanner, the parser or the semantic analyzer.


\subsubsection{Lexical errors}
If the scanner finds an error, it outputs a message and returns a special error token to the parser.
Following suggestions, identifiers with more than 32 characters are no longer considered an error
and only output a warning if they are found, which does not block compilation.
The scanner recognizes the following errors (Table \ref{tab:scanner-errors}).

\begin{table}[h]
\centering
\captionsetup{width=0.7\textwidth}
\caption{Lexical errors and their associated tokens. The tokens are returned
to the parser and the program continues analyzing the rest of the text.}
\label{tab:scanner-errors}
\begin{tabular} {c c}
\hline
Error type & Token type \\
\hline
Unrecognized token & \texttt{UNRECOGNIZED\_TOKEN} \\
Malformed character constant & \texttt{INVALID\_CHAR\_CONST} \\
\hline
\end{tabular}
\end{table}

\subsubsection{Syntax errors}
Syntax errors are thrown from the parser when it is unable to complete a shift or reduce
operation. To allow the parser to continue reading the input, the built in \texttt{error}
token is included in some of the grammar's rules. Particularly, we have included
such rules in order to capture: 

\begin{itemize}
\item  Malformed blocks of code (the parser continues after the next semicolon or enclosing
curly brace)
\begin{lstlisting}
declaration
	: ...
	| error ';'
	| error compound_statement
	;

\end{lstlisting}

\item  Malformed statements (the parser continues after the next semicolon)
\begin{lstlisting}
statement
	: ...
	| error ';'
\end{lstlisting}

\item  Malformed expressions (the parser continues after the next semicolon or enclosing
parenthesis)
\begin{lstlisting}
primary_expression
	: ...
	| error
\end{lstlisting}

\item  Malformed argument lists (the parser continues after the next enclosing parenthesis)
\begin{lstlisting}
function_definition
	: ...
	| type IDENTIFIER '(' error ')'
\end{lstlisting}
\end{itemize}


\subsubsection{Semantic errors}
Semantic errors are thrown from the semantic analyzer when there are constructs
in the input text that violate the language's semantics as described in the Section
\ref{section:semantics}. These errors do not abort the compilation process before
the parser has finished, as most of the semantic analysis is performed
simultaneously with the parsing. Semantic errors are handled by the functions
defined in the file ``\texttt{actions.c}''.

\paragraph{Re-declaration of variable:} This error is issued if a variable is
declared or initialized more than once with the same name in a given scope.
It is \textbf{not} an error to declare of initialize a variable with the same
name of one that exists in a parent scope, but the new variable will shadow the
old one, that is, the variable in the parent scope will be inaccessible. % issue warning in the future

\paragraph{Redefinition of function:} This error is issued if a function is
defined more than once with the same name \textbf{and} the same argument types.
This language aims to implement polymorphic functions, therefore, defining functions
with the same name and different argument types is allowed.

\paragraph{Variable use without declaration:} This error is issued if a variable is
used before being declared or initialized either in the current scope or any of its
parent scopes up to the global scope. Variable declaration or initialization must
always precede its use.

\paragraph{Function call without definition:} This error is issued if a function
is called but no function with that name has been defined. Functions must always
be defined before they are used.

\paragraph{Function call with invalid argument types:} This error is issued if a
previously defined function is called with arguments types that are not present
in any of the previous definitions of that function.

\paragraph{Incompatible types:} This error is issued if there is an expression
with operands of incompatible types, as described on Section \ref{section:types}.

\subsubsection{Remarks about error handling}
These strategies allows us to display a reasonable amount of errors before the program is unable
to proceed and terminates. In fact the parser should be able to finish (albeit not correctly)
the syntax tree even when errors are found in several situations.
The error messages are displayed as they are found, and since the
scanner returns a token that is never used by the parser, lexical errors also generate syntax
errors. To avoid encumbering the output when errors are found, only error messages and the symbol
table are displayed on invalid input.

